---
title: "Ottbikes forecasting notebook"
output: html_notebook
---

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(tsibble)
library(feasts)
library(forecast)
library(GGally)
library(zoo)
library(ggfortify)
```


```{r}
bikes.ts <- readRDS("../data/ottbike_counters.ts.RDS") %>%
  mutate(location = as.factor(location))
levels(bikes.ts$location)

locations <- c("Adawé", "Alexandra", "Canal Ritz", "Colonel By","Laurier/Bay",  "Laurier/Lyon", "Laurier/Metcalfe", "OTrain/Bayview", "OTrain/Gladstone", "OTrain/Young", "Ottawa River", "Portage", "Somerset")
```



### Imputing missing values?

-fill gaps just adds NA to gaps within the min-max dates

```{r}
bikes.ts %>% 
  filter(!is.na(count)) %>%
  as_tsibble(index = Date, key = location) %>% 
  ggplot(aes(x=Date, y=location, color=log(count+1))) + geom_tile() +
  scale_colour_viridis_c() + theme_light()

bikes.ts %>% 
  filter(!is.na(count)) %>% 
  as_tsibble(index = Date, key = location) %>% fill_gaps() %>%
  filter(as.character(location) %in% locations[c(2,3,4,11)]) %>%
  # mutate(count = na.approx(count)) %>%
  ggplot(aes(x=Date, y=location, color=log(count+1))) + geom_tile() +
  scale_colour_viridis_c() + theme_light()
```

- the naive na.approx() method does poorly considering the strong seasonal trends that should be fairly predictable? Imputation could produce a more acceptable simulated version.

```{r}
bikes.ts %>% 
  filter(!is.na(count)) %>% 
  as_tsibble(index = Date, key = location) %>% fill_gaps() %>%
  filter(as.character(location) %in% locations[c(2,3,4,11)]) %>%
  mutate(count = zoo::na.approx(count)) %>%
  ggplot(aes(x=Date, y=location, color=count)) + geom_tile() +
  scale_colour_viridis_c() + theme_light()
```



### The missing values are really messing up the overall trend

```{r}
stl.comps <- bikes.ts %>% 
  filter(!is.na(count)) %>% as_tsibble(index = Date, key = location) %>%
  fill_gaps() %>%
  filter(as.character(location) %in% locations[c(2)]) %>%
  mutate(count = na.approx(count)) %>%
  model(STL(count ~ season(window=13),  robust = TRUE)) %>%
  components()

stl.comps %>%
  autoplot(alpha=0.5, colour = 'red') + theme(legend.position = 'bottom')



stl.comps2 <- bikes.ts %>% 
  filter(!is.na(count)) %>% 
  as_tsibble(index = Date, key = location) %>%
  filter(Date >= ymd("2011-01-01") & Date < ymd("2018-01-01")) %>%
  fill_gaps() %>%
  filter(as.character(location) %in% locations[c(2)]) %>%
  mutate(count = na.approx(count)) %>%
  model(STL(count ~ season(window=13),  robust = TRUE)) %>%
  components()

stl.comps2 %>%
  autoplot(alpha=0.5, colour = 'blue') + theme(legend.position = 'bottom')
```



```{r}
library(patchwork)

stl.comps2 <- stl.comps2 %>%
  mutate(location = "filtered")
stl.comps3 <- stl.comps %>%
  bind_rows(stl.comps2) %>%
  mutate(location = as.factor(location))

p1 <- stl.comps3  %>%
  autoplot(count, alpha=0.5, facets = TRUE) + 
  theme(legend.position = 'none', axis.title.x = element_blank(),
        axis.text.x = element_blank(), axis.title.y = element_text(angle=0))
p2 <- stl.comps3  %>%
  autoplot(trend, alpha=0.5, facets = TRUE) + 
  theme(legend.position = 'none', axis.title.x = element_blank(),
        axis.text.x = element_blank(), axis.title.y = element_text(angle=0))

p3 <- stl.comps3  %>%
  autoplot(season_year, alpha=0.5, facets = TRUE) + 
  theme(legend.position = 'none', axis.title.x = element_blank(),
        axis.text.x = element_blank(), axis.title.y = element_text(angle=0))

p4 <- stl.comps3  %>%
  autoplot(season_week, alpha=0.5, facets = TRUE) + 
  theme(legend.position = 'none', axis.title.x = element_blank(),
        axis.text.x = element_blank(), axis.title.y = element_text(angle=0))  

p5 <- stl.comps3  %>%
  autoplot(remainder, alpha=0.5, facets = TRUE) + 
  theme(legend.position = 'none', axis.title.x = element_blank(),
        axis.text.x = element_blank(), axis.title.y = element_text(angle=0))

p6 <- stl.comps3  %>%
  autoplot(season_adjust, alpha=0.5, facets = TRUE) + 
  theme(legend.position = 'none', axis.title.x = element_blank(),
        axis.text.x = element_blank(), axis.title.y = element_text(angle=0))
```


```{r}
p1/p2/p3/p4/p5/p6 & theme_light() + theme(legend.position = 'none', axis.title.x = element_blank(),
        axis.text.x = element_blank(), axis.title.y = element_text(angle=0))
```














### Features list from *feasts* package:


```{r}
bikes_features <- bikes.ts %>%
  features(count, feature_set(pkgs="feasts"))
bikes_features
```

- missing some coefficients
- need a way to present stats data in charts


```{r}
# bikes_features %>%
#   select_at(vars(contains("season"), location)) %>%
#   mutate(
#     seasonal_peak_week = glue::glue("Q{seasonal_peak_week+1}"),
#     seasonal_trough_week = glue::glue("Q{seasonal_trough_week+1}"),
#   ) %>%
#   GGally::ggpairs(mapping = aes(colour=location))
```

